---
title: "Tutorial script for multiple-task joint modeling with the PMwG sampler"
output:
  html_document:
    df_print: paged
---
```{r}
knitr::opts_chunk$set( 
  eval = FALSE,
  echo = TRUE
)
```

## Overview
This code constructs a joint model of two simulated behavioural datasets.
It can be extended to N tasks and multiple data sources per subject 
(e.g., EEG + choice-RT + fMRI). The code and variable names in this 
script are intended to be as general as possible to allow easy 
modification and to let users 'plug in' their own experimental designs 
and models as straightforwardly as possible.

The script starts by creating an experimental design matrix (an N x M data
frame representing N trials of M experimental factors). The design matrix
encodes the features of the experimental design on each trial and is 
useful for simulating and plotting data, and computing variables based on 
your particular experimental design. As such, the most important objects 
in the script - the data frames called `data`, `design`, and `sims`- all 
have the same form as the design matrix and can thus each be passed 
interchangeably to the likelihood and sampling functions for different 
purposes (e.g., fitting to data, simulating out of a model, generating 
hypothetical data). 

At the heart of this code is the likelihood function. The likelihood
function iterates over each trial in the design and computes either 
1) the likelihood of the current observation in `data` given the current 
parameter values, or 2) a simulated observation generated from the model
given the current parameters and cell of the experimental design.

The bulk of the work when building a new model is writing the likelihood
function. It is easiest to first write (and debug) the 'stand alone' 
likelihood of the model as applied to a single trial (row) in the data. 
From there you can easily iterate over trials using a for loop. 
Likelihood functions for many designs can be made faster by 'vectorizing' 
over observations that share a cell in the design matrix (e.g., with a 
single call to `dLBA()`, `dnorm()`, etc.). However, the loop method has the 
advantage of being more intuitive and easier to debug in the initial 
stages of writing a model despite its longer compute time. 

Constructing a joint likelihood (e.g., the likelihood of N models of N 
data sets) simply involves putting each model's likelihood loop inside 
the same likelihood function and passing that function a named list of 
data frames (e.g., the list `data` containing data frames `data$task1`, 
`data$task2`, `data$task3`, etc.). The returned likelihood is then just 
the grand sum of each models individual summed likelihood. Functions for 
simulating out of the models are computed in the same manner as the 
single case and are returned as a list of data frames (one for each 
task/model) at the end. 

Running the Particle Metropolis within Gibbs (PMwG) sampler is 
straightforward - you simply pass the `data` (or `sims`) object, the 
likelihood function, and a vector of parameter names as arguments. 
See [the online tutorial](https://newcastlecl.github.io/samplerDoc/) 
for further information about the sampler. 

When working with a single model, the PMwG sampler accounts for 
covariation among parameters hierarchically, via a multivariate
normal distribution with mean vector containing the hyper-level
means of each parameter and variance-covariance matrix representing
shared variance among parameters. When working with multiple models 
simultaneously (i.e., joint modeling), the mean vector is simply
the union (concatenation) of the individual model's mean vectors 
(e.g., `all.means <- c(m1.means, m2.means))`. The covariance matrix 
is similarly constructed based on this extended means vector (thus 
variance shared among different tasks is accounted for hierarchically
under an overarching multivariate normal).
 
__TO DO:__ Add convenience functions for plotting parameter samples, model
  fits, posterior predictives/parameter recovery/model exploration, 
  heatmap of hyper var-covar matrix, etc.

## Setup environment 

#### Clear workspace
```{r}
rm(list=ls())
```

#### Set working directory
```{r}
getwd()
main_dir <- c("/Users/rboag/Dropbox/My Mac (Russell’s MacBook Pro)/Documents/PMwG")
setwd(file.path(main_dir))
```

#### Create subdirectory structure 
This will not overwrite if the directory already exists.
```{r}
sub_dirs <- list("samples", "plots", "data")
lapply(sub_dirs, function (x) dir.create(file.path(main_dir, x), showWarnings = FALSE))

```

#### Load packages
```{r}
library("tidyverse")
library("rtdists")
library("pmwg")
```

#### Set seed for random number reproducibility
```{r}
set.seed(101)
```

#### Create some empty storage objects
```{r}
design <- list()
data <- list()
p_names <- list()
p_vector <- list()
sims <- list()
```

We will fill these in with important bits and pieces needed by the 
likelihood and sampling functions as we go. These data objects are 
the 'building blocks' of how we will pass data to and from the likelihood
and sampling functions. 

We will mostly work with lists of data frames (or vectors) because this 
allows us to generalize the code to an arbitrary number of 
tasks/models/data sources when doing joint modeling. 
 
To this end, all the important data objects we will work with conform to
the `tidyverse` principles of tidy data and can thus be used as is with 
the `purrr` and `broom` libraries for performing the same computation on 
many objects. This is especially useful for modeling, plotting, and 
summarizing joint models and data spanning multiple tasks and subjects.

## Setup Task 1 
Before building the combined joint model, we will first build up and test 
all the necessary pieces (e.g., design, data, likelihood) for each 
task/model separately. We will combine them in the final step once we are
confident the individual pieces each work correctly. 

#### Make experimental design matrix for Task 1
Here we create an N x M data frame of our experimental design. 
Each row represents one observation or trial and each column is a factor 
whose levels code the experimental conditions for each trial. 
For example, the following generates a design matrix for a simple 2 x 2 
(`stim` x `cond1`) speed-accuracy trade-off experiment with column factors
`subject`, `stim`, and `cond1`. Fully factorial designs are particularly
easy to create by passing vectors of subjects and factor levels to 
`expand_grid()`. More complex and unbalanced designs, and designs 
containing randomization or sequential effects, may need to be 
constructed manually (e.g., by assigning to cells the output of 
`sample()` or `ifelse()`). 
```{r}
n_subs_t1 <- 5  # n subjects
n_obs_t1 <- 40  # n observations (trials) per subject per design cell
```

#### Factor levels
```{r}
levels_stim_t1 <- c("s1", "s2")
levels_cond1_t1 <- c("spd", "acc")
```

#### Create design matrix
Here we use `expand_grid()` to generate a factorial combination of all 
subjects, stimuli, and speed-accuracy cue conditions. Note that your 
empirical data should also be in a similarly tidy format with column 
names and factor levels that match those in your likelihood function. 
```{r}
design$t1 <- expand_grid(subject = factor(rep(1:n_subs_t1, each = n_obs_t1)), 
                         stim = factor(levels_stim_t1),
                         cond1 = factor(levels_cond1_t1)
                         )
design$t1
```

#### Extract parameter names
Now we want to create a named vector containing all the unique estimated
parameters in our design. First we extract the factor levels over which 
the parameters will vary. This creates a unique text string label for 
each parameter in the design. The likelihood and sampling functions use
this to lookup parameters and pull out values. 
```{r}
pars_a_t1 <- str_c("t1.a", unique(design$t1$cond1), sep = ".")
pars_v_t1 <- str_c("t1.v", unique(design$t1$cond1), sep = ".")
pars_t0_t1 <- str_c("t1.t0", sep = ".")
```

#### Create a vector of parameter names
This is particularly important when passing multiple similar models to 
the sampler. Here we let a and v vary over condition and keep t0 fixed 
(i.e., we don't pass it any factor levels via `unique()`). 
```{r}
p_names$t1 <- c(pars_a_t1, pars_v_t1, pars_t0_t1)
p_names$t1
```

#### Check the number of parameters makes sense
```{r}
length(grep("t1.a", p_names$t1))
length(grep("t1.v", p_names$t1))
length(grep("t1.t0", p_names$t1))
```

#### Set mean parameter values to simulate from (or to use as priors)
Note that the values of unestimated constants are set inside the 
likelihood function.
```{r}
p_vector$t1 <- c(2.0, 3.0,
                 1.0, 1.5,
                 0.2)
names(p_vector$t1) <- p_names$t1
p_vector$t1
```

Looks good. Now we can construct the likelihood function for Task 1.

## Likelihood function Task 1
Although the following function is quite long, it is very modular. It is
made of three self-explanatory code chunks: 

1. Check inputs
2. Likelihood loops
3. Prepare output

Adding a model to the function involves adding its likelihood loop to 
the 'Likelihood loops' chunk and appending the result to the `data` and
`out` objects before computing the summed log-likelihood.

Notice that the relevant data is indexed by name (e.g., `data$t1$rt`). 
This makes adding additional data sources easy: you simply append the 
new source to the data list and give it a name (e.g., `data$t1_EEG`, 
`data$t2_BOLD`). Refer to it by that name when writing the likelihood.

#### Likelihood function for drift diffusion model of Task 1
```{r}
ll_t1_ddm <- function(x, data, sample = FALSE) {
  # This function takes the following inputs:
  # 1. x = the (log of) a named vector of starting parameters.
  # 2. data = an experimental data frame or the experimental 
  #    design matrix created above if simulating data (i.e., 
  #    if sample = TRUE).
  #
  # This function returns the following outputs:
  # 1. The sum of the log-likelihoods (if sample = FALSE)
  # 2. A data frame of simulated data (if sample = TRUE)
  
  # Check inputs ------------------------------------------------------------
  
  # Undo the log transform
  x <- exp(x) 
  # Perform parameter checks
  if (any(data$t1$rt < x["t1.t0"])) {
    return(-1e10)
  }
  
  if (sample) {
    # Create output objects for simulated data
    data$t1$rt <- rep(NA, nrow(data$t1))
    data$t1$response <- rep(NA, nrow(data$t1))
  } else {
    # Create output objects for likelihoods
    out <- list()
    out$t1 <- numeric(nrow(data$t1))
  }
  
  # Likelihood loops --------------------------------------------------------
  
  # Task 1 loop -------------------------------------------------------------
  
  # Now we iterate over each trial in the data (or design matrix) and 
  # either simulate an outcome or calculate the likelihood given the 
  # current parameters
  for (i in 1:nrow(data$t1)) {
    # Extract the parameters that correspond to the current trial
    t1.a = x[paste("t1.a", data$t1$cond1[i], sep = ".")]
    t1.v = x[paste("t1.v", data$t1$cond1[i], sep = ".")]
    t1.t0 = x[paste("t1.t0")]
    
    if (sample) {
      # Simulate a trial from the model using the current parameters
      tmp <- rdiffusion(n = 1,
                        a = t1.a,
                        v = t1.v,
                        t0 = t1.t0
      )
      data$t1$rt[i] <- tmp$rt
      data$t1$response[i] <- tmp$response
    } else {
      # Calculate the likelihood (density) for the current parameters
      out$t1[i] <- ddiffusion(rt = data$t1$rt[i],
                              response = data$t1$response[i],
                              a = t1.a,
                              v = t1.v,
                              t0 = t1.t0
      )
    }
  }
  
  # Task 2 loop -------------------------------------------------------------
  
  # Here is where you insert additional likelihood loops for joint modeling.
  # Remember to append results produced here to `data` and `out` below.
  
  # Prepare output ----------------------------------------------------------
  
  if (sample) {
    # Output simulated data
    data$t1$response <- factor(data$t1$response)
    return(data$t1) # Remove index if returning multiple
  } else {
    # Remove bad likelihoods
    out <- out$t1
    bad <- (out < 1e-10) | (!is.finite(out))
    out[bad] <- 1e-10
    # Output summed log-likelihood
    out <- sum(log(out))
    return(out)
  }
}
```

## Test likelihood function
Here we check that out likelihood function returns the expected items 
(i.e., a simulated data)

#### Simulate data from the design matrix
```{r}
sims$t1 <- ll_t1_ddm(x = log(p_vector$t1), 
                     data = design, 
                     sample = TRUE)
sims$t1
```

#### Return likelihood
```{r}
ll_t1_ddm(x = log(p_vector$t1), 
          data = sims, 
          sample = FALSE)
```

## Explore data

#### Load data
Here we would usually read in and tidy up our empirical data to get it 
into a similar format as the data frame we just simulated and stored in 
`sims`. For now let's just use the simulated data as our `data`. 
```{r}
data <- sims
data$t1
```

#### Cell counts
```{r}
table(data$t1$stim, data$t1$cond1)
```

#### Mean RT
```{r}
data$t1 %>% 
  group_by(stim, cond1) %>% 
  summarise(rt = mean(rt))
```

#### RT histogram
```{r}
data$t1 %>%
  ggplot(aes(x = rt, fill = stim)) +
  geom_histogram(stat = "bin") +
  facet_wrap(. ~ cond1) +
  scale_x_continuous(limits = c(0, 3)) +
  theme_minimal()
```

#### __TO DO:__ Add any other useful data checks here...

Now we can do the same for Task 2...

## Setup Task 2 

#### Make experimental design matrix for Task 2
```{r}
n_subs_t2 <- 5
n_obs_t2 <- 40
levels_stim_t2 <- c("s1", "s2")
levels_cond1_t2 <- c("spd", "acc")
design$t2 <- expand_grid(subject = factor(rep(1:n_subs_t2, each = n_obs_t2)), 
                         stim = factor(levels_stim_t2),
                         cond1 = factor(levels_cond1_t2)
)
design$t2
```

#### Extract parameter names
```{r}
pars_a_t2 <- str_c("t2.a", unique(design$t2$cond1), sep = ".")
pars_v_t2 <- str_c("t2.v", unique(design$t2$cond1), sep = ".")
pars_t0_t2 <- str_c("t2.t0", sep = ".")
```

#### Create a vector of parameter names
This is particularly important when passing multiple similar models to 
the sampler. Here we let a and v vary over condition and keep t0 fixed 
(i.e., we don't pass it any factor levels via `unique()`). 
```{r}
p_names$t2 <- c(pars_a_t2, pars_v_t2, pars_t0_t2)
p_names$t2
```

#### Check the number of parameters makes sense
```{r}
length(grep("t2.a", p_names$t2))
length(grep("t2.v", p_names$t2))
length(grep("t2.t0", p_names$t2))
```

#### Set mean parameter values to simulate from (or to use as priors)
Note that the values of unestimated constants are set inside the 
likelihood function.
Here we will just set them to half of model 1's values.
```{r}
p_vector$t2 <- c(2.0, 3.0,
                 1.0, 1.5,
                 0.2)/2
names(p_vector$t2) <- p_names$t2
p_vector$t2
```

## Likelihood function Task 2 

#### Likelihood function for drift diffusion model of Task 2
```{r}
ll_t2_ddm <- function(x, data, sample = FALSE) {
  # This function takes the following inputs:
  # 1. x = the (log of) a named vector of starting parameters.
  # 2. data = an experimental data frame or the experimental 
  #    design matrix created above if simulating data (i.e., 
  #    if sample = TRUE).
  #
  # This function returns the following outputs:
  # 1. The sum of the log-likelihoods (if sample = FALSE)
  # 2. A data frame of simulated data (if sample = TRUE)
  
  # Check inputs ------------------------------------------------------------
  
  # Undo the log transform
  x <- exp(x) 
  # Perform parameter checks
  if (any(data$t2$rt < x["t2.t0"])) {
    return(-1e10)
  }
  
  if (sample) {
    # Create output objects for simulated data
    data$t2$rt <- rep(NA, nrow(data$t2))
    data$t2$response <- rep(NA, nrow(data$t2))
  } else {
    # Create output objects for likelihoods
    out <- list()
    out$t2 <- numeric(nrow(data$t2))
  }
  
  # Likelihood loops --------------------------------------------------------
  
  # Task 2 loop -------------------------------------------------------------
  
  # Now we iterate over each trial in the data (or design matrix) and 
  # either simulate an outcome or calculate the likelihood given the 
  # current parameters
  for (i in 1:nrow(data$t2)) {
    # Extract the parameters that correspond to the current trial
    t2.a = x[paste("t2.a", data$t2$cond1[i], sep = ".")]
    t2.v = x[paste("t2.v", data$t2$cond1[i], sep = ".")]
    t2.t0 = x[paste("t2.t0")]
    
    if (sample) {
      # Simulate a trial from the model using the current parameters
      tmp <- rdiffusion(n = 1,
                        a = t2.a,
                        v = t2.v,
                        t0 = t2.t0
      )
      data$t2$rt[i] <- tmp$rt
      data$t2$response[i] <- tmp$response
    } else {
      # Calculate the likelihood (density) for the current parameters
      out$t2[i] <- ddiffusion(rt = data$t2$rt[i],
                              response = data$t2$response[i],
                              a = t2.a,
                              v = t2.v,
                              t0 = t2.t0
      )
    }
  }
  
  # Prepare output ----------------------------------------------------------
  
  if (sample) {
    # Output simulated data
    data$t2$response <- factor(data$t2$response)
    return(data$t2) # Remove index if returning multiple
  } else {
    # Remove bad likelihoods
    out <- out$t2
    bad <- (out < 1e-10) | (!is.finite(out))
    out[bad] <- 1e-10
    # Output summed log-likelihood
    out <- sum(log(out))
    return(out)
  }
}
```

## Test likelihood function 

#### Simulate data from the design matrix
```{r}
sims$t2 <- ll_t2_ddm(x = log(p_vector$t2), 
                     data = design, 
                     sample = TRUE)
sims$t2
```

#### Return likelihood
```{r}
ll_t2_ddm(x = log(p_vector$t2), 
          data = sims, 
          sample = FALSE)
```

## Explore data

#### Load data
Here we would usually read in and tidy up our empirical data to get it 
into a similar format as the data frame we just simulated and stored in 
`sims`. For now let's just use the simulated data as our `data`. 
```{r}
data <- sims
data$t2
```

#### Cell counts
```{r}
table(data$t2$stim, data$t2$cond1)
```

#### Mean RT
```{r}
data$t2 %>% 
  group_by(stim, cond1) %>% 
  summarise(rt = mean(rt))
```

#### RT histogram
```{r}
data$t2 %>%
  ggplot(aes(x = rt, fill = stim)) +
  geom_histogram(stat = "bin") +
  facet_wrap(. ~ cond1) +
  scale_x_continuous(limits = c(0, 3)) +
  theme_minimal()
```


## Create joint model
Now we are ready to combine the two models into a single likelihood. First let's make a combined parameter vector containing all parameters.

#### Make combined parameter vector
```{r}
p_vector <- c(p_vector$t1, 
              p_vector$t2)
p_vector
```

## Setup for sampling

#### Get the names of the parameters we want to sample
```{r}
pars <- names(p_vector)
pars
```

#### Set priors for the mean and variance of hyper-level multivariate normal 
```{r}
priors <- list(
  theta_mu_mean = rep(0, length(pars)),
  theta_mu_var = diag(rep(1, length(pars)))
)
priors
```

## Joint likelihood function

#### Likelihood function for joint drift diffusion model of Tasks 1 & 2
```{r}
ll_joint_ddm <- function(x, data, sample = FALSE) {
  # This function takes the following inputs:
  # 1. x = the (log of) a named vector of starting parameters
  # 2. data = a named list of data frames or experimental design  
  #    matrices if simulating data (i.e., if sample = TRUE)
  #
  # This function returns the following outputs:
  # 1. The sum of the log-likelihoods (if sample = FALSE)
  # 2. A list of data frames of simulated data (if sample = TRUE)
  
  # Check inputs ------------------------------------------------------------
  
  # Undo the log transform
  x <- exp(x) 
  # Perform parameter checks 
  # Task 1 checks
  if (any(data$t1$rt < x["t1.t0"])) {
    return(-1e10)
  }
  # Task 2 checks
  if (any(data$t2$rt < x["t2.t0"])) {
    return(-1e10)
  }
  
  if (sample) {
    # Create output objects for simulated data
    # Task 1 sim outputs
    data$t1$rt <- rep(NA, nrow(data$t1))
    data$t1$response <- rep(NA, nrow(data$t1))
    # Task 2 sim outputs
    data$t2$rt <- rep(NA, nrow(data$t2))
    data$t2$response <- rep(NA, nrow(data$t2))
  } else {
    # Create output objects for likelihoods
    out <- list()
    out$t1 <- numeric(nrow(data$t1))
    out$t2 <- numeric(nrow(data$t2))
  }
  
  # Likelihood loops --------------------------------------------------------
  
  # Task 1 loop -------------------------------------------------------------
  
  # Now we iterate over each trial in the data (or design matrix) and 
  # either simulate an outcome or calculate the likelihood given the 
  # current parameters
  for (i in 1:nrow(data$t1)) {
    # Extract the parameters that correspond to the current trial
    t1.a = x[paste("t1.a", data$t1$cond1[i], sep = ".")]
    t1.v = x[paste("t1.v", data$t1$cond1[i], sep = ".")]
    t1.t0 = x[paste("t1.t0")]
    
    if (sample) {
      # Simulate a trial from the model using the current parameters
      tmp_t1 <- rdiffusion(n = 1,
                           a = t1.a,
                           v = t1.v,
                           t0 = t1.t0
      )
      data$t1$rt[i] <- tmp_t1$rt
      data$t1$response[i] <- tmp_t1$response
    } else {
      # Calculate the likelihood (density) given the current parameters
      out$t1[i] <- ddiffusion(rt = data$t1$rt[i],
                              response = data$t1$response[i],
                              a = t1.a,
                              v = t1.v,
                              t0 = t1.t0
      )
    }
  }
  
  # Task 2 loop -------------------------------------------------------------
  
  # Again we iterate over each trial and either simulate an outcome
  # or calculate the likelihood given the current parameters
  for (i in 1:nrow(data$t2)) {
    # Extract the parameters that correspond to the current trial
    t2.a = x[paste("t2.a", data$t2$cond1[i], sep = ".")]
    t2.v = x[paste("t2.v", data$t2$cond1[i], sep = ".")]
    t2.t0 = x[paste("t2.t0")]
    
    if (sample) {
      # Simulate a trial from the model using the current parameters
      tmp_t2 <- rdiffusion(n = 1,
                           a = t2.a,
                           v = t2.v,
                           t0 = t2.t0
      )
      data$t2$rt[i] <- tmp_t2$rt
      data$t2$response[i] <- tmp_t2$response
    } else {
      # Calculate the likelihood (density) given the current parameters
      out$t2[i] <- ddiffusion(rt = data$t2$rt[i],
                              response = data$t2$response[i],
                              a = t2.a,
                              v = t2.v,
                              t0 = t2.t0
      )
    }
  }
  
  # Prepare output ----------------------------------------------------------
  
  if (sample) {
    # Output simulated data
    data$t1$response <- factor(data$t1$response)
    data$t2$response <- factor(data$t2$response)
    return(data)
  } else {
    # Remove bad likelihoods
    out_all <- c(out$t1, out$t2)
    bad <- (out_all < 1e-10) | (!is.finite(out_all))
    out_all[bad] <- 1e-10
    # Output summed log-likelihood
    # print(paste("ll t1: ", sum(log(out$t1))))
    # print(paste("ll t2: ", sum(log(out$t2))))
    out_all <- sum(log(out_all))
    return(out_all)
  }
}
```

## Test likelihood function

#### Simulate data from the design matrix
```{r}
sims <- ll_joint_ddm(x = lp_vector, data = data, sample = TRUE)
sims
```

#### Return likelihood
```{r}
ll_joint_ddm(x = lp_vector, data = sims, sample = FALSE)

```

## Setup sampling 

#### Initialize sampler
```{r}
sampler <- pmwgs(
  data = data,
  pars = pars,
  prior = priors,
  ll_func = ll_joint_ddm
)
```

#### Set start points (optional)
```{r}
start_points <- list(
  mu = log(p_vector),
  sig2 = diag(rep(.01, length(pars)))
)
```

#### Initialize start points
```{r}
sampler <- init(
  sampler, 
  start_mu = start_points$mu,
  start_sig = start_points$sig2
)
```

## Run sampling

#### Stage 1: Burn-in
```{r}
burned <- run_stage(
  sampler, 
  stage = "burn",
  iter = 500,
  particles = 2000,
  epsilon = .5, 
  n_cores = 10
)
```

#### Stage 2: Adaptation
```{r}
adapted <- run_stage(
  burned, 
  stage = "adapt", 
  iter = 10000, 
  particles = 2000,
  epsilon = .5,
  n_cores = 10
)
```

#### Stage 3: Sampling
```{r}
sampled <- run_stage(
  adapted, 
  stage = "sample",
  iter = 1000, 
  particles = 200,
  epsilon = .5,
  n_cores = 10
)
```

#### Save to samples directory
```{r}
save(sampled, file = "samples/samples_ddm_combined_test.RData")
```
